---
title: 'Assignment 3: Model comparison'
author: "Marton Kovacs"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

In this lab assignment you are going to work with (simulated) data related to perioperative pain and its psychological and hormonal predictors. In the assignment you will assess the added benefit of including some psychological and hormonal predictors to the already established demographic predictors of pain.

In this assignment you will set up a hierarchical regression model to predict postoperative pain after wisdom tooth surgery. 

# Research problem

The amount of pain experienced around and after surgeries are highly variable between and within individuals. In order to improve surgical pain management regimens we need to understand what influences pain around surgical procedures and predict the amount of pain an individual will experience.

Your first study in this area is related to assessing the influence of trait and state psychological measures on pain, and to see whether taking into account these variables can improve our understanding of postoperative pain.

# Procedures and measures

Use the data file called ‘assignment_3_dataset’, from the 'data/' folder.

You have collected data from 160 adults who were scheduled to undergo surgical extraction of the third mandibular molar (wisdom tooth surgery). Patients filled out a form in the waiting room before their surgery. The form contained questions about their sex, age, and weight, and psychological questionnaires assessing anxiety, pain catastrophizing, and mindfulness (see descriptions below). You also got blood samples and saliva samples from participants in the waiting room 5 minutes before their operations to determine the serum (a component of the blood) and salivary cortisol levels of participants. Participants were contacted 5 hours after the surgery to see how much pain they were experiencing. The __level of pain__ at that moment was recorded using a numerical rating scale using a __scale of 0 to 10__, where 0 means “no pain” and 10 means “worst pain I can imagine”. 

__The State Trait Anxiety Inventory:__ T measures trait anxiety on a scale of 20 to 80, higher scores mean higher anxiety. Anxiety has been found in many studies to positively correlate with the level of pain experienced. This is __variable STAI_trait__ in the dataset.

__The Pain Catastrophizing Scale__ measures the extent of pain catastrophizing, which is characterized by a tendency to magnify the threat value of a pain stimulus and to feel helpless in the presence of pain, as well as by a relative inability to prevent or inhibit pain-related thoughts in anticipation of, during, or following a painful event. The total score on this scale ranges from 0 to 52, higher scores mean higher catastrophizing. Pain catastrophizing is one of the well-established predictors of clinical pain. This is __variable pain_cat__ in the dataset.

__The Mindful Attention Awareness Scale (MAAS)__ measures dispositional mindfulness, which may be described as a tendency to turn attention to present-moment experiences in an open, non-judgmental way. The MAAS total score ranges from 1 to 6 (an average of the item scores), with higher scores representing higher dispositional mindfulness. Trait mindfulness has been theorized to serve as a protective factor against pain, as the individual would be more objective about their pain experience and tend to associate less discomfort, despair, and hopelessness to the pain-related sensations. This is __variable mindfulness__ in the dataset.

__Cortisol__ is a stress hormone associated with acute and chronic stress. Cortisol levels are thought to be positively associated with pain experience. Cortisol can be __measured from both blood and the saliva__, although, serum cortisol is often regarded in medical research as more reliably related to stress (serum is a component of the blood plasma). These are __variables cortisol_serum__, and __cortisol_saliva__ in the dataset.

# Research question

Previous studies and meta-analyses showed that age and sex are often predictors of pain (age is negatively associated with pain, while sex is a predictor more dependent on the type of the procedure). You would like to determine the extent to which taking into account psychological and hormonal variables aside from the already used demographic variables would improve our understanding of postoperative pain.

To answer this research question you will __need to compare two models__ (with a hierarchical regression). The __simpler model__ should contain __age and sex as predictors of pain__, while the __more complex model__ should contain the __predictors: age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures__. Notice that the predictors used in the simpler model are a subset of the predictors used in more complex model. __You will have to do model comparison to assess whether substantial new information was gained about pain in the more complex model compared to the simpler model.__  

# What to report

As usual, before you can interpret your model, you will need to run data and model diagnostics. First, check the variables included in the more complex model (age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures as predictors, and pain as an outcome) for __coding errors__, and the model itself for __influential outliers__ (for example using Cook’s distance). Furthermore, check the final model to see if the __assumptions of linear regression hold true__, that is, __normality__ (of the residuals), __linearity__ (of the relationship), __homogeneity of variance__ (also called homoscedasticity) and that there is no excess __multicollinearity__ (“uncorrelated predictors” in Navarro’s words). If you find anything amiss during these checks, make the appropriate decision or correction and report your findings and actions in your report. 

__Note:__ If you do any changes, such as exclude cases, or exclude predictors from the model, you will have to re-run the above checks for your final data and model.

Report the results of the simpler model and the more complex model. For both models you should report the model test statistics (adj.R2, F, df, and p value). Also, report the statistics describing the coefficients of the predictors in a table format (unstandardized regression coefficients and 95% confidence intervals, standardized regression coefficients (B and Beta values), and p values).

Write up the regression equation of the more complex model in the form of 𝑌 = 𝑏0 + 𝑏1 ∗ X1 + 𝑏2 ∗ X2 +…+ bn * Xn, in which you use the actual regression coefficients of your models. (b0 stands for the intercept and b1, b2 … bn stand for the model coefficients for each of the predictors, and X1, X2, … Xn denote the predictors).

Compare the two models in terms of how much variance they explain of pain’s variability in the sample. Report Akaike information criterion (AIC) for both models and the F test statistic and p value of the likelihood ratio test comparing the two models.

# What to discuss

In your discussion of the findings, briefly interpret the results of the above analyses, and indicate whether you think that anything was gained by including the psychological and hormone measures in the model.

# Solution

## Read the data

Read the dataset used in this assignment. Pay attention to the extension of the datafile. Load the needed packages as well.

```{r}
library(knitr)
library(readxl)

# Supress warnings
options(warn = -1)

# Reading the dataset
assignment_3_dataset <- read_excel("~/GitHub/public_r_data_analysis_2021_fall/data/assignment_3_dataset.xlsx")

# Load the needed packages for completing the exercises
library(skimr) 
library(faraway)
library(car)
library(QuantPsyc)
library(tidyverse)
```

## Data and model diagnostics 
### Data diagnostics
#### Descriptives of the variables

Run an exploratory data analysis (EDA) to investigate the dataset.

```{r}
# Looking at the raw data
glimpse(assignment_3_dataset)
assignment_3_dataset

# Look at the mean and distribution of each variable
assignment_3_dataset %>% select(pain, sex, age, STAI_trait, pain_cat, cortisol_serum, cortisol_saliva, mindfulness) %>% skim()

# Look at the correlations between the variables
assignment_3_dataset %>% 
  select(pain, age, STAI_trait, pain_cat, cortisol_serum, cortisol_saliva, mindfulness) %>%
  cor()
```

**From the exploratory data analyses we can see that no value is missing from the dataset. The mean values of the different variables are around the middle of the possible range of values. The standard deviation is smaller in case of the cortisol measures and the mindfulness questionnaire, while larger in case of the pain, age, STAI and pain catastrophizing. From the minimum and maximum values it can be seen that there is a coding error, as the pain measure was from 0 to 10, but one participant indicated a pain level of 50. By looking at the dataset it can also be seen that in the sex variable one subject was coded as "woman" instead of "female". The correlations showed that between most of the variables there is a small or medium correlation, but between cortisol_saliva and cortisol_serum the correlation is very high (above 0.9).**

#### Correct coding errors

If you find values in the dataset during the EDA, that are not correct based on the provided descriptions of the variables of the dataset please correct them here.

```{r}
# Correct the coding errors
assignment_3_dataset[142, 2] <- 5
assignment_3_dataset[35, 3] <- "female"
```

### Model diagnostics
#### Build the more complex model

In order to test the more complex model for outliers and to test the assumptions first build the model.

```{r}
# Building the complex model
complex_model  <- lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_saliva + cortisol_serum, data=assignment_3_dataset)
summary(complex_model)
```

#### Checking for influential outliers

Check for outlier values in the model.

```{r}
# Adding variables to the original dataset which will help to detect outliers
assignment_3_dataset$residuals<-resid(complex_model)
assignment_3_dataset$standardized.residuals<- rstandard(complex_model)
assignment_3_dataset$studentized.residuals<-rstudent(complex_model)
assignment_3_dataset$cooks.distance<-cooks.distance(complex_model)
assignment_3_dataset$dfbeta<-dfbeta(complex_model)
assignment_3_dataset$dffit<-dffits(complex_model)
assignment_3_dataset$leverage<-hatvalues(complex_model)
assignment_3_dataset$covariance.ratios<-covratio(complex_model)

# Saving the dataset with the new variables
write.table(assignment_3_dataset, "Assignment3_dataset With Diagnostics.dat", sep = "\t", row.names
= FALSE)

# Creating a new variable which shows whether each standardized residual is larger or smaller than 2 (called large residuals), then, summing the number of such large residuals
assignment_3_dataset$large.residual <- assignment_3_dataset$standardized.residuals > 2 | assignment_3_dataset$standardized.residuals < -2
sum(assignment_3_dataset$large.residual) 

 # Looking at which cases has standardized residual larger than 2 or -2 
assignment_3_dataset[assignment_3_dataset$large.residual,c("ID", "pain", "age", "sex", "STAI_trait", "pain_cat", "mindfulness", "cortisol_saliva", "cortisol_serum", "standardized.residuals")]

# Examining further the cases with large residuals, in terms of Cook's distance, leverage and covariance ratios
assignment_3_dataset[assignment_3_dataset$large.residual, c("cooks.distance", "leverage", "covariance.ratios")] 
```

**We expected to see that 95% of the values has a stardardized residual <+/-2. In our sample of 160 this would be 8. It can be seen that 11 residuals are larger than 2 or -2 which is 3 more than it is allowed. Only 1 case is above 2.5, but this is allowed as we expect that 1% of the data will have a standardized residual greater than 2.5 or -2.5. No standardized residual is above 3. None of the cases has a Cook's distance greater than 1, which means that no cases has an undue influence on the model. The average leverage is k+1/n=7+1/160=0.05, all leverage values should be 3 times the average leverage (0.15), which criteria is fulfilled in each case with a large residual, as we can see from the output. The boundaries of the covariance ratios are the following: CVR1 > 1+(3(k+1)/n)=1+(3*7+1)/160)=1.15; CVR2 < 1-(3(k+1)/n)=1-(3*7+1)/160)=0.85; case 105, 107, and 150 are outside this limit.**

#### Checking assumptions

Check the normality assumption.

```{r}
# Plotting the distribution of the residuals to see whether the residuals have a normal distribution
ggplot(assignment_3_dataset, aes(x = residuals)) +
  geom_histogram(binwidth = 0.25, color = "white") +
  labs(x = "Residual")
```

**The histogram shows a fairly normal distribution of the residuals.**

Check the linearity assumption.

```{r}
# Checking linearity
plot(complex_model, 1) 
```

**Although there is a small curve for the fitted line on the plot, it does not seem to violate the assumption of linearity significantly.**

Check the homoscedasticty assumption (homogeneity of variance).

```{r}
# Checking homoscedasticity 
plot(complex_model, 3)
```

**The plot shows that the homoscedasticity assumption is met.**

Check the multicollinearity assumption.

(VIF above 5), or a VIF threshold of 3 is recommended in this paper: http://onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2009.00001.x/full

Some info about VIF: 
https://statisticalhorizons.com/multicollinearity
http://blog.minitab.com/blog/understanding-statistics/handling-multicollinearity-in-regression-analysis

```{r}
# Calculating the VIF for the complex model
vif(complex_model)

# Calculating tolerance
1/vif(complex_model) 
```

**The VIF for the cortisol_serum and cortisol_saliva variables was above 3, while tolerance was below 0.2, which indicates a problem with multicollinearity. The solution is dropping the variable cortisol_saliva from the model (because cortisol_serum is the more reliable measure of cortisol, it is better to drop the saliva measure).**

### Making decision based on model diagnostics

If based on the assumption tests you decide to drop a predictor variable you should do that here. Create your updated model.

```{r}
# Drop cortisol_saliva from the model and create the updated model
complex_model_upd  <- lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum, data=assignment_3_dataset)
summary(complex_model_upd)
```

#### Checking outliers of the updated model

```{r}
# Adding variables to a newly created dataframe for the updated model, which will help to detect outliers
assignment_3_dataset_updmodel<-assignment_3_dataset
assignment_3_dataset_updmodel$residuals<-resid(complex_model_upd)
assignment_3_dataset_updmodel$standardized.residuals<- rstandard(complex_model_upd)
assignment_3_dataset_updmodel$studentized.residuals<-rstudent(complex_model_upd)
assignment_3_dataset_updmodel$cooks.distance<-cooks.distance(complex_model_upd)
assignment_3_dataset_updmodel$dfbeta<-dfbeta(complex_model_upd)
assignment_3_dataset_updmodel$dffit<-dffits(complex_model_upd)
assignment_3_dataset_updmodel$leverage<-hatvalues(complex_model_upd)
assignment_3_dataset_updmodel$covariance.ratios<-covratio(complex_model_upd)

# Saving the dataset with the new variables
write.table(assignment_3_dataset_updmodel, "Assignment3_dataset With Diagnostics of the updated complex model.dat", sep = "\t", row.names
= FALSE)

# Creating a new variable which shows whether each standardized residual is larger or smaller than 2 (called large residuals), then, summing the number of such large residuals
assignment_3_dataset_updmodel$large.residual <- assignment_3_dataset_updmodel$standardized.residuals > 2 | assignment_3_dataset_updmodel$standardized.residuals < -2
sum(assignment_3_dataset_updmodel$large.residual) # We see that 10 residuals are larger than 2 or -2 which is 2 more than it is allowed. Only 1 case is above 2.5, but this is allowed as we expect that 1% of the data will have a standardized residual greater than 2.5 or -2.5. No standardized residuals are above 3.

# Looking at which cases has standardized residual larger than 2 or -2 
assignment_3_dataset_updmodel[assignment_3_dataset_updmodel$large.residual,c("ID", "pain", "age", "sex", "STAI_trait", "pain_cat", "mindfulness", "cortisol_serum", "standardized.residuals")] 

# Examining further the cases with large residuals, in terms of Cook's distance, leverage and covariance ratios
assignment_3_dataset_updmodel[assignment_3_dataset_updmodel$large.residual , c("cooks.distance", "leverage", "covariance.ratios")]
```

**We expect to see that 95% of the values has a stardardized residual <+/-2. In our sample of 160 this would be 8. It can be seen that 10 residuals are larger than 2 or -2 which is 2 more than it is allowed. No standardized residuals are above 2.5. None of the cases has Cook's distance greater than 1, which means that no cases has an undue influence on the model. The average leverage is k+1/n=6+1/160=0.043, all leverage values should be 3 times the average leverage (0.13), which criteria is fulfilled in each case with a large residual, as we can see from the output. The boundaries of the covariance ratios are the following: CVR1 > 1+(3(k+1)/n)=1+(3*(6+1)/160)=1.13; CVR2 < 1-(3(k+1)/n)=1-(3*6+1)/160)=0.87; only case 107 and 150 are outside this limit.**

#### Checking assumptions of the updated model

Normality assumption

```{r}
# Plotting the distribution of the residuals in the updated complex model to see whether the residuals have a normal distribution
ggplot(assignment_3_dataset_updmodel, aes(x = residuals)) +
  geom_histogram(binwidth = 0.25, color = "white") +
  labs(x = "Residual")
```

**The histogram shows a fairly normal distribution of the residuals.**

Linearity assumption

```{r}
# Checking linearity
plot(complex_model_upd, 1) # Although there is a small curve for the fitted line on the plot, it does not seem to violate the assumption of linearity significantly

```

**Although there is a small curve for the fitted line on the plot, it does not seem to violate the assumption of linearity significantly.**

Homoscedasticty assumption (homogeneity of variance)

```{r}
# Checking homoscedasticity 
plot(complex_model_upd, 3) # The plot shows that the homoscedasticity assumption is met
```

**The plot shows that the homoscedasticity assumption is met.**

Multicollinearity assumption

```{r}
# Calculating the VIF for the updated complex model
vif(complex_model_upd) 

# Calculating tolerance
1/vif(complex_model_upd) 
```

**All VIF and tolerance values are acceptable, there is no problem with multicollinearity.**

## Model comparison

Create the simple model and get the results of the model that needs to be reported based on the What to report section.

```{r}
# Creating the simple model 
simple_model  <- lm(pain ~ age+ sex, data=assignment_3_dataset)
```

Create the more complex model based on the results of the model diagnostics. Also, get the results that needs to be reported based on the What to report section.

```{r}
# The final model is the updated model
summary(complex_model_upd)
summary(simple_model)

## Obtaining Beta values and confidence intervals
lm.beta(complex_model_upd)
lm.beta(simple_model)
confint(complex_model_upd)
confint(simple_model)
```

### Table of results

#### Model Test Statistics
Models | Adjusted R-squared | F | df | p
-----|-----|-----|-----|-----
Simple Model | 0,076 | 7,555 | 157 | <0,001
Complex Model | 0,326 | 13,85 | 153 | <0,001

#### Predictors in simple model
Predictors | b (unstandardized coefficient) | 95% CI lower bound | 95% CI upper bound | Beta (standardized coefficient) | p
-----|-----|-----|-----|-----|-----
Intercept | 8,49 | 6,63 | 10,36 | - | <0,001
Age | -0,09 | -0,13 | -0,04 | -0,3 | <0,001
Sexmale | 0,1 | -0,36 | 0,56 | NA | 0,66

#### Predictors in complex model
Predictors | b (unstandardized coefficient) | 95% CI lower bound | 95% CI upper bound | Beta (standardized coefficient) | p
-----|-----|-----|-----|-----|-----
Intercept | 1,99 | -1,35 | 5,32 | - | 0,24
Age | -0,04 | -0,08 | 0,01 | -0,12 | 0,12
Sexmale | 0,31 | -0,1 | 0,73 | NA | 0,14
STAI_trait | -0,01 | -0,06 | 0,04 | -0,04 | 0,67
Pain_cat | 0,08 | 0,03 | 0,14 | 0,27 | 0,003
Mindfulness | -0,15 | -0,39 | 0,01 | -0,09 | 0,24
Cortisol_serum | 0,53 | 0,29 | 0,78 | 0,35 | <0,001

**Regression equation for the complex model: 
pain = b0 + b1 X age + b2 X sexmale + b3 X STAI_trait + b4 X pain_cat + b5 X mindfulness + b6 X cortisol_serum=1.98 -0.036 X age + 0.31 X sexmale -0.011 X STAI_trait + 0.084 X pain_cat -0.15 X mindfulness + 0.53 X cortisol_serum**

Compare the two models.

```{r}
# Comparing the two model
anova(simple_model, complex_model_upd) # 
AIC(complex_model_upd, simple_model)
```

Discussion: 
In your discussion of the findings, briefly interpret the results of the above analyses, and indicate whether you think that anything was gained by including the psychological and hormone measures in the model.

**Both regression models are significant, which means they give a better prediction of pain than the mean of pain. The complex model explains 33% of the variation in pain, while the simple model explains 7%. The complex model significantly improved the fit of the model to the data compared to simple_model, F(4, 153) = 15.599, p < .001.
AIC is 533.36 for the complex model, and 580.09 for the simple model. Because AIC is lower in the complex model, it has a better fit. Based on these analyses it seems that the complex model is better, thus, the model that includes both the psychological and hormone measures is better than the model that only contained age and sex to predict perioperative pain after wisdom tooth surgery. When looking at the individual predictors in the complex model, we can see that when controlling for all the other predictors, only pain catastrophization and the serum cortisol level is significant. These two measures are the most important in predicting perioperative pain.**

